Assault Rate Trends Within and Around the University of Toronto: A Ten-Year Analysis

### 2024-01-24 outputs/paper/paper.qmd
commiting all work done in paper.qmd thus far.

## Load packages
library(tidyverse)
library(knitr)
library(ggplot2)
library(readr)

@@

#read in cleaned data #
surrounding_neighbourhoods <- read.csv(here::here("outputs/data/analysis_data.csv"))
show_col_types = FALSE
```

@@

###Source and Data Collection
This dataset includes the Crime Data by Neighbourhood. Counts are available for Assault, Auto Theft, Break and Enter, Robbery, Theft Over, Homicide and Shooting & Firearm Discharges. Data also includes the crime rate per 100,000 population calculated using the population estimates provided by Environics Analytics. Comparing to crime count, crime rate provides a fairer comparison of the crime over time by taking into account the change in population in the region. The data was last refreshed on Jan 11, 2024 and is refreshed anually. 
I conducted the first step of basic data cleaning to eliminate all counts and rates not applicable to this paper, so all I was left with was assault rates per neighbourhood for each year. I then refined the nieghbourhoods to include only those that encompass the University of Toronto and then its surrounding areas. The city of Toronto recognizes 158 unique Neighbourhoods and of the 158, 13 have been included in the cleaned data.

@@

##Results

>>>>

## Load packages
library(tidyverse)
library(knitr)
library(ggplot2)
library(readr)
library(dplyr)
library(tidyr)
library(stringr)

@@

#read in cleaned data #
surrounding_neighbourhoods <- read.csv(here::here("outputs/data/analysis_data.csv"))
show_col_types = FALSE

#read in cleaned data #
all_neighbourhoods <- read.csv(here::here("outputs/data/discussion_data.csv"))
show_col_types = FALSE
```

@@

###Source and Data Collection
This dataset includes the Crime Data by Neighbourhood. Counts are available for Assault, Auto Theft, Break and Enter, Robbery, Theft Over, Homicide and Shooting & Firearm Discharges. Data also includes the crime rate per 100,000 population calculated using the population estimates provided by Environics Analytics. Comparing to assault count, assault rate provides a fairer comparison of the crime over time by taking into account the change in population in the region. The data was last refreshed on Jan 11, 2024 and is refreshed annually. 
I conducted the first step of basic data cleaning to eliminate all counts and rates not applicable to this paper, so all I was left with was assault rates per neighbourhood for each year. I then refined the nieghbourhoods to include only those that encompass the University of Toronto and then its surrounding areas. The city of Toronto recognizes 158 unique neighbourhoods and of the 158, 13 have been included in the cleaned data.

@@

##Results
###Assault Rates per University of Toronto Neighbourhood 
Over the past decade, University of Toronto's encompassing neighbourhoods have exhibited unique trends in assault rates. "Kensington-Chinatown" shows an erratic pattern, peaking around 2018 followed by a sharp decline in 2019 and a subsequent rise again. "The Annex" demonstrates a generally consistent, however, slight upward trend. Both "Bay-Cloverhill" and "University" show quite inconsistent patterns with several fluctuations over the ten year time span. What is most notable is Universities steady increase from 2020 on. The Annex also shows an increase from 2020 on, though not as steep as University. (\@ref(fig-assaults_per_year_uoft)) shows the assault rates per year for each University of Toronto neighbourhood.
```{r}
#| message: false
#| echo: false
#| label: fig-assaults_per_year_uoft
#| fig-cap: Assault Rates per year in University of Toronto Neighbourhoods (2014-2023)

long_data <- cleaned_data |>
  gather(key = "Year", value = "Assault_Rate", -Neighbourhood) |>
  mutate(Year = as.numeric(str_replace(Year, "ASSAULT_RATE_", "")))

# Plot the graph
ggplot(long_data, aes(x = Year, y = Assault_Rate, group = Neighbourhood, color = Neighbourhood)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Assault Rates in University of Toronto 
      Neighbourhoods (2014-2023)",
    x = "Year",
    y = "Assault Rate",
    color = "Neighbourhood"
  ) +
  ylim(0, 5000) # Adjust these values based on your data's range

```

###Assault Rates per Neighbourhood in the Surrounding Areas
Assault rates in the neighbourhoods surrounding the University of Toronto Neighbourhoods reveals some distinct patterns and fluctuations in the data.  Several neighbourhoods exhibit fairly stable trends with minor year-to-year changes including "Yonge-St.Clair", "Wychwood", "West Queen West, "Palmerston-Little Italy", "Case Loma" and "Trinity-Bellwoods". "Dovercourt Village" shows a slight upward trend over the decade. "Church-Wellesley" demonstrate quite a sporadic trend with several jumps and falls. From 2014-2020 "Wellington Place" shows a steep downward trend, rising from 2020-2022, however, then showing a steep decline again from 2022-2023. Several neighbourhoods exhibit quite low assault rates over the entire decade including "Yonge-St. Clair", "Casa Loma", "Palmerston-Little Italy", and "Wychwood". For a full picture of assault rates from 2014 to 2023 across various neighbourhoods surrounding the University of Toronto, please refer to (\@ref(fig-assaults_per_year_surounding_neighbourhoods)).
```{r}
#| message: false
#| echo: false
#| label: fig-assaults_per_year_surounding_neighbourhoods
#| fig-cap: Assault Rates in Surrounding Neighbourhoods

long_data <- surrounding_neighbourhoods |>
  gather(key = "Year", value = "Assault_Rate", -Neighbourhood) |>
  mutate(Year = as.numeric(str_extract(Year, "\\d{4}")))

# Checking for NA values in Year because when I was building my graph orignally the output was a graph with nothing plotted on it even though everything else looked good the Year column is producing NA values
sum(is.na(long_data$Year))

# Plot the graph
ggplot(long_data, aes(x = Year, y = Assault_Rate, group = Neighbourhood, color = Neighbourhood)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(
    title = "   Assault Rates in University of Toronto's 
Surrounding Neighbourhoods (2014-2023)",
    x = "Year",
    y = "Assault Rate",
    color = "Neighbourhood"
  ) +
  scale_x_continuous(breaks = seq(from = 2014, to = 2023, by = 2)) # I added this line because it specifies the break points for the x-axis, which in this context represents the years. This line is telling ggplot2 to place axis ticks at every integer value between 2014 and 2023. I had to add this because the years were not showing up there was something wrong at the x-axis was coming up blank and therefore there was no datapoints in the graph

```
#Discussion
The data allows for a comparative analysis of assault rates across various neighborhoods in and around the University of Toronto from 2014 to 2023. By comparing the assault trends in the university's immediate vicinity with those of the surrounding neighborhoods, I aim to uncover underlying patterns, draw attention to areas of concern, and spark discussions around potential causative factors.

The data presents a counterintuitive narrative, revealing that the University of Toronto, a Canadian fortress of education and innovation, experiences higher assault rates on average compared to its neighbouring communities. This finding challenges common perceptions of campus environments as insulated and safe. People often believe that a university is a safe bubble, away from the problems of the outside world, especially at a prestigious university focused on cutting edge learning and progress. However, that is not what the data portrays. (See\@ref(fig-assaults_per_year_per_neighbourhoods))

```{r assault-rates-line-plot, fig.width=12, fig.height=8, echo=FALSE}
#| message: false
#| echo: false
#| label: fig-assaults_per_year_per_neighbourhoods
#| fig-cap: Assault Rates per year Across All Neighbourhoods (2014-2023)

# Your data preparation code
long_data <- all_neighbourhoods |>
  gather(key = "Year", value = "Assault_Rate", -Neighbourhood) |>
  mutate(Year = as.numeric(str_extract(Year, "\\d{4}")))

# Plot the graph with facets and without a legend
ggplot(long_data, aes(x = Year, y = Assault_Rate, group = Neighbourhood, color = Neighbourhood)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Assault Rates in University of Toronto's Surrounding Neighbourhoods (2014-2023)",
    x = "Year",
    y = "Assault Rate"
  ) +
  scale_x_continuous(breaks = seq(from = 2014, to = 2023, by = 1)) +
  facet_wrap(~ Neighbourhood, scales = "free_x") +
  theme(
    legend.position = "none", # This line removes the legend
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  )

```
